Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-05-26 analysis/paper_draft4.Rmd
split paper and supplement into separate files (paper_draft4.Rmd and paper_draft4supp.Rmd)

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
header-includes: 
  - \usepackage{amsmath}
  - \usepackage{subfig}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, sex, or membership in certain other cognizable classes. Here, we extend a Bayesian approach for detecting such illegal strike bias by showing how to incorporate historical data on an attorney's use of peremptory strikes in past cases. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. Using simulations, we show how the choice of the power prior's discounting parameter influences bias detection (how likely the credible interval for the bias parameter excludes zero), depending on the degree of incompatibility between current and historical trial data. Finally, we extend this approach with a prototype software application that lawyers could use to detect strike bias in real time during jury-selection. We illustrate this application's use with real historical strike data from a convenience sample of cases from one court.

@@

In the United States, individuals selected for jury-service appear in court as scheduled and are questioned by the parties' attorneys and the trial judge. During this process, a prospective juror, if not excused by the trial judge for cause, may still be dismissed if a party's attorney uses one of their limited number of peremptory challenges against them. By asserting a peremptory challenge, a party can declare a prospective juror ineligible ("strikes" that juror) for a seat on the jury without the burden of explaining why.

Since _Batson v. Kentucky_ [-@batson1986], a party violates the Equal Protection Clause of the US Constitution by using peremptory challenges if motivated by the prospective jurors' race, ethnicity, sex, or certain other cognizable classes. The party bringing a _Batson_ challenge bears the burden of proving such illegal bias is more likely than not to be true [@LaFave2022, $\S$ 22.3(d)].

@@

To evaluate the proposed method, we conducted a simulation study using Stan `r rstan::stan_version()`, a Hamiltonian Monte Carlo engine for Bayesian inference by way of `r R.version$version.string` and RStan [@rstan]. These simulations primarily show how much the size of the historical data affects our ability to detect bias and how much the power prior discounted the weight of the historical data. As introduced above, $b > 0$ represents a prosecutor's bias against a prospective juror within the cognizable class, while $b = 0$ denotes no bias. For simplicity, we assume the defense attorney has no bias.

@@

The choice of credible interval also affects bias detection. For example, with a 90\% credible interval, when $b_{curr}=2$, the bias-detection rate is lower than with an 80\% credible interval; when $b_{curr}=0$, that detection rate is higher. The tradeoff is lower accuracy. With a 90\% credible interval, when the bias parameter is 2, the model is less accurate in detecting bias. When the bias parameter is 3, however, the detection rate is still high. For details on simulation results for 90% and 80% credible intervals, see the Appendix.

@@

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app with `r R.version$version.string` and the shiny package [@R; @shiny]. Unlike the simulations, we built the app using Rcpp [@Rcpp] and the Metropolis-Hasting algorithm to sample $b$ from the posterior distribution. As in the simulations, a normal distribution was assumed for the distribution of $b$. The length of the Markov chain was 110000 and the first 10,000 iterations were dropped as burn-in. No thinning was performed, as the correlation was weak and convergence occurred rapidly. We checked convergence with both traceplots and the Gelman-Rubin statistic [@Gelman1992]. Computation times vary from 0.76 to 1.2 seconds, depending in the amount of data used (quicker with current strike data only _versus_ current plus historical strike data).

To show how lawyers might use this app in real cases in real time during jury selection, we loaded this app with actual strike data from a convenience sample of attorneys who appeared during jury selection in criminal cases in the federal district court for Connecticut during fiscal years 2013 through 2017. Using publicly-available court docket sheets for each case, we matched particular lawyers to these strikes and later assigned them aliases using the charlatan package [@charlatan]. For details on this dataset, see the Appendix.

@@

In our illustration, we select the names of the prosecutor and defense attorney from their respective pull-down menus; and leave the weight setting to "Equal". The density plots update accordingly (Figure \@ref(fig:figapp1)(b)). Now the credible intervals clearly exclude zero. Thus, we have a credible basis to infer bias against racial-minority prospective jurors in how these attorney use their peremptory challenges in the present case. In the Appendix, we present additional illustrations wherein we use the app to detect gender bias and where we assign less weight to the historical strike data.

@@

# CRediT Author Statement.

@@

# Appendix

Here, we extend section \@ref(simulation-results) by reporting our simulation results for 90% and 80% credible intervals.  We also extend the discussion in section \@ref(prototype) by discussing data collection and presenting additional illustrations of how a lawyer might use the prototype app.

## Bias Detection with 90% and 80% Credible Intervals

The choice of credible interval can affect bias detection as we define it, because the larger the credible interval one chooses, the greater the chance that zero falls within it, and vice-versa.  Here, we report simulation results for bias detection for 90% intervals for trials with fifteen, ten, and six strikes per side (Figures \@ref(fig:figbd9015), \@ref(fig:figbd9010), and \@ref(fig:figbd906), respectively), and the same for 80% credible intervals (Figures \@ref(fig:figbd8015), \@ref(fig:figbd8010), and \@ref(fig:figbd806)).

```{r figbd9015, out.width = "95%", fig.cap = "Bias Detection Rates, 90 pct credible interval, 15 strike scenario", fig.align='center'}

load_path <- here::here("figures","pp15_90CI.png")
knitr::include_graphics(load_path)

```{r figbd9010, out.width = "95%", fig.cap = "Bias Detection Rates (90 pct credible interval, 10 strikes scenario)", fig.align='center'}

load_path <- here::here("figures","pp10_90CI.png")
knitr::include_graphics(load_path)

```{r figbd906, out.width = "95%", fig.cap = "Bias Detection Rates based on 90 pct credible interval for 6 strike scenario", fig.align='center'}

load_path <- here::here("figures","pp6_90CI.png")
knitr::include_graphics(load_path)

```{r figbd8015, out.width = "95%", fig.cap = "Bias Detection Rates, 80 pct credible interval, 15 strike scenario", fig.align='center'}

load_path <- here::here("figures","pp15_80CI.png")
knitr::include_graphics(load_path)

```{r figbd8010, out.width = "95%", fig.cap = "Bias Detection Rates (80 pct credible interval, 10 strikes scenario)", fig.align='center'}

load_path <- here::here("figures","pp10_80CI.png")
knitr::include_graphics(load_path)

```{r figbd806, out.width = "95%", fig.cap = "Bias Detection Rates based on 80 pct credible interval for 6 strike scenario", fig.align='center'}

load_path <- here::here("figures","pp6_80CI.png")
knitr::include_graphics(load_path)

## Prototype Software Application

### Data Collection

To collect the historical strike data for the prototype software application, one of us filed a request with the U.S. District Court of Connecticut based on 28 U.S.C. $\S$ 1868.  Thereafter, we received copies of certain jury-selection records associated with twenty-nine criminal cases in that court during this period. These included strike sheets that indicated the identification number of prospective jurors who were struck by peremptory challenge, the order in which they were struck, and which side (prosecutor or defense) struck which juror. Such records also included a tally of answers to juror questionnaires that asked each prospective juror to report their race and gender.

These records, however, often did not indicate the identity of the attorneys exercising the strikes. While the standard forms included a signature line for the attorney, many were left blank or filled with illegible signatures.  Accordingly, we turned to the publicly-available docket sheets for each case for the names of the lawyers who appeared in the case on behalf of the prosecution (the US Attorney's office) or the criminal defendant(s) on the date(s) of jury selection. Where only a single attorney represented a party during jury selection, it was easy to attribute the pattern of strikes to that attorney. Where multiple lawyers appeared for one side, we attributed to each of them that side's pattern of peremptory challenges in that case. In such cases, neither the jury-selection documents nor the docket sheets indicated any hierarchy among multiple lawyers or any other basis to attribute strikes to only one attorney among them. After generating a dataset based on these documents, we kept only strikes where a criminal defendant was represented by an attorney. Finally, to de-identify this dataset, we excluded defendant names and substituted fictitious names for the attorneys using the charlatan package [@charlatan].

### Additional Illustrations

Suppose the same scenario discussed in section \@ref(prototype), except now the strike tally indicates the pattern of strikes by these attorneys against female prospective jurors. If so, we select "gender" as the cognizable class, and the density plots update accordingly (Figure \@ref(fig:figapp3)(left)).

```{r figapp3, out.width='50%', fig.cap = "Screenshots of R-Shiny application show density plots for gender bias of prosecutor and defense attorney based on historical and current strike data. Vertical dotted lines depict 95 pct. credible interval.", fig.show = 'hold', fig.ncol = 2, fig.subchap=c("Equal weight","Minimal weight") }

load_path3 <- here::here("figures","batson_app_screenshots", "batson_app_screenshot3.png")
load_path4 <- here::here("figures","batson_app_screenshots", "batson_app_screenshot4.png")

#par(mfrow=c(2,1))

knitr::include_graphics(load_path3)
knitr::include_graphics(load_path4)

Here, most of the prosecutor's density curve of both prior and posterior are to the left of zero, indicating the possible bias against male jurors. However, because the 95\% credible interval includes zero, inferring such bias from the strike data alone is unjustified. Finally, if we assign minimal weight to the historical strike data ($a = 0.1$), the density plot updates accordingly (Figure \@ref(fig:figapp3)(right)). Both the prior and posterior density become flatter and the credible interval plainly includes zero.

>>>>

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
header-includes: 
  - \usepackage{amsmath}
  - \usepackage{subfig}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  United States law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, sex, or membership in certain other cognizable classes. Here, we extend a Bayesian approach for detecting such illegal strike bias by showing how to incorporate historical data on an attorney's use of peremptory strikes in past cases. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. Using simulations, we show how the choice of the power prior's discounting parameter influences bias detection (how likely the credible interval for the bias parameter excludes zero), depending on the degree of incompatibility between current and historical trial data. Finally, we extend this approach with a prototype software application that lawyers could use to detect strike bias in real time during jury-selection. We illustrate this application's use with real historical strike data from a convenience sample of cases from one court.

@@

In the United States, individuals selected for jury service appear in court as scheduled and are questioned by the parties' attorneys and the trial judge. During this process, a prospective juror, if not excused by the trial judge for cause, may still be dismissed if a party's attorney uses one of their limited number of peremptory challenges against them. By asserting a peremptory challenge, a party can declare a prospective juror ineligible ("strikes" that juror) for a seat on the jury without the burden of explaining why.

Since _Batson v. Kentucky_ [-@batson1986], a party violates the Equal Protection Clause of the United States Constitution by using peremptory challenges if motivated by the prospective jurors' race, ethnicity, sex, or certain other cognizable classes. The party bringing a _Batson_ challenge bears the burden of proving such illegal bias is more likely than not to be true [@LaFave2022, $\S$ 22.3(d)].

@@

To evaluate the proposed method, we conducted a simulation study using Stan 2.261, a Hamiltonian Monte Carlo engine for Bayesian inference by way of R version 4.2.2 and RStan and RStan [@rstan]. These simulations primarily show how much the size of the historical data affects our ability to detect bias and how much the power prior discounted the weight of the historical data. As introduced above, $b > 0$ represents a prosecutor's bias against a prospective juror within the cognizable class, while $b = 0$ denotes no bias. For simplicity, we assume the defense attorney has no bias.

@@

The choice of credible interval also affects bias detection. For example, with a 90\% credible interval, when $b_{curr}=2$, the bias-detection rate is lower than with an 80\% credible interval; when $b_{curr}=0$, that detection rate is higher. The tradeoff is lower accuracy. With a 90\% credible interval, when the bias parameter is 2, the model is less accurate in detecting bias. When the bias parameter is 3, however, the detection rate is still high. For details on simulation results for 90% and 80% credible intervals, see the Supplementary Material.

@@

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app with `r R.version$version.string` and the shiny package [@R; @shiny]. Unlike the simulations, we built the app using Rcpp [@Rcpp] and the Metropolis-Hasting algorithm to sample $b$ from the posterior distribution. As in the simulations, a normal distribution was assumed for the distribution of $b$. The length of the Markov chain was 110,000 and the first 10,000 iterations were dropped as burn-in. No thinning was performed, as the correlation was weak and convergence occurred rapidly. We checked convergence with both traceplots and the Gelman-Rubin statistic [@Gelman1992]. Computation times vary from 0.76 to 1.2 seconds, depending in the amount of data used (quicker with current strike data only _versus_ current plus historical strike data).

To show how lawyers might use this app in real cases in real time during jury selection, we loaded this app with actual strike data from a convenience sample of attorneys who appeared during jury selection in criminal cases in the federal district court for Connecticut during fiscal years 2013 through 2017. Using publicly-available court docket sheets for each case, we matched particular lawyers to these strikes and later assigned them aliases using the charlatan package [@charlatan]. For details on this dataset, see the Supplementary Material.

@@

In our illustration, we select the names of the prosecutor and defense attorney from their respective pull-down menus; and leave the weight setting to "Equal". The density plots update accordingly (Figure \@ref(fig:figapp1)(b)). Now the credible intervals clearly exclude zero. Thus, we have a credible basis to infer bias against racial-minority prospective jurors in how these attorney use their peremptory challenges in the present case. In the Supplementary Material, we present additional illustrations wherein we use the app to detect gender bias and where we assign less weight to the historical strike data.

@@

# CRediT Author Statement

@@
